---
title: "Red Wine Exploration"
author : "Ernestas Žemaitis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

#Packages to load
library(ggplot2)
library(GGally)
library(scales)
library(lattice)
library(MASS)
library(memisc)

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('/home/ernestas/MEGAsync/Udacity/R_projects/Project/')
pf <- read.csv('wineQualityReds.csv')
pf <- na.omit(pf)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Show column headers
names(pf)
```

Here is a data on each colum about value distribution between each wine cases.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Statistical data about each column
summary(pf)
```

What is clear from this table is that quality ratings only varies only from 3 to
8.2nd - density variable isn't vary much which might not have effect for quality
ratings. 3rd - We only have 1599 whine cases which is not arebig number for the
analysis. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot quality rating histogram
ggplot(aes(x=quality), data=pf) +
  geom_histogram(color = 'black', binwidth = 1) +
  scale_x_continuous("Quality rating", breaks = seq(1,10,1)) +
  scale_y_continuous("Count") +
  ggtitle('Quality rating distribution') +
  theme(plot.title = element_text(hjust = 0.5))
```

Most vines are valued in between 5 and 6 quality rating. However, there are way
lower number of cases where wine ratings are below 5 and higher than 6. Quality
ratings are distributed symetrically in between 3 and 8.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=fixed.acidity), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.5)
```

Fixed acidity distribution is right skewed with a peak of around 7. Values are
in range 4.6 and 15.9.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=volatile.acidity), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.1) 
```

Violatile acidity distribution is right skewed with a peak of 0.6 . Volatile
acidity is varying from 0.12 to 1.58.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=citric.acid), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.05)
```

Citric acidity values varies between 0 and 1. Distribution is similar to right
skewed shape with highest number of values at 0 and 0.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=residual.sugar), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.25)
```

Residual sugar varies from 0.9 all the way to 16. However most values are in
ragne 1.9 to 2.6. This distribution is right skewed.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=chlorides), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.01)
```

Chlorides distribution is Right skewed. Most values are at 0.79 and varies from 
0.012

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=free.sulfur.dioxide), data=pf) +
  geom_histogram(color = 'black', binwidth = 2)
```

Distribution is right skewed. Most values are about 5 and it varies from 1 to 72.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=total.sulfur.dioxide), data=pf) +
  geom_histogram(color = 'black', binwidth = 5)
```

Total sulfur dioxide is also right skewed. These values varies a lot: from 6 to
289. Most values are low ones - from 6 to 70.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=density), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.001)
```

Density distribution is symetrical.However wine density does not vary a lot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=pH), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.05)
```

pH values stays in acid range. It varies from 2.7 to 4.1. Values distributed
symetrically.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=sulphates), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.05)
```

Sulphate values distributed in right skewed shape.Values ranges from 0.33 to 2.
Most values are between 9 and 9.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot histogram
ggplot(aes(x=alcohol), data=pf) +
  geom_histogram(color = 'black', binwidth = 0.25)
```

Looks like alcohol values are also distributed in right skewed shape. Values
ranges 8.49 to 14.90. Looks like most values are between 9 and 10.

# Univariate Analysis

### What is the structure of your dataset?

This tidy data set contains 1,599 red wines with 11 variables on the 
chemical properties of the of the Portuguese "Vinho Verde" red wine. 
At least 3 wine experts rated the quality of each wine, providing a rating 
between 0 (very bad) and 10 (very excellent).According to these ratings we will 
try find what properties of wine that make wine get the highest ratings. Here we
can see all collumns.

### What is/are the main feature(s) of interest in your dataset?

Chemical features of wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

None of the atributes are stand out. However alcohol, fixed acidity, 
violatile acidity, citric acidity, free & total sulfur dioxides are interest
of mine because these atributes varies in wider range.

### Did you create any new variables from existing variables in the dataset?

I haven't created yet.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

We checked for misisng values, however none of them were identified.

# Bivariate Plots Section

Because nothing stands out, we will try to plot every variable with quality
atrribute.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=volatile.acidity), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Volatile acidity") +
  ggtitle('Wine quality by volatile acidity') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

As we can see violatile acidity tend to be lower, when quality ranking is going
higher.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=fixed.acidity), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Fixed acidity") +
  ggtitle('Wine quality by fixed acidity') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Mean and median values are almost the same accross all quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=citric.acid), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Citric acidity") +
  ggtitle('Wine quality by citric acidity') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Citric acidity tend to be higher in higher quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=residual.sugar), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Residual sugar") +
  ggtitle('Wine quality by residual sugar') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Looking to quality ratings, residual sugar stays almost the same accros the
ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=chlorides), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Chlorides") +
  ggtitle('Wine quality by chlorides') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Chlorides also stays the same.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=density), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Density") +
  ggtitle('Wine quality by density') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Density has a slight decrease, however it's values varies in very small numbers. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=total.sulfur.dioxide), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Total sulfur dioxide") +
  ggtitle('Wine quality by total sulfur dioxide') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Total sulfur dioxide remains almost the same except of slight peak at rating
of 5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=pH), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("pH") +
  ggtitle('Wine quality by pH') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

pH is tend to be slighty more acid

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=sulphates), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Sulphates") +
  ggtitle('Wine quality by sulphates') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Sulphates has a small increase in higher quality rating wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot scatter plot with mean and median lines.
ggplot(aes(x=quality, y=alcohol), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Alcohol") +
  ggtitle('Wine quality by citric alcohol') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

Average and median values of alchocol tend to be almost the same from ratings 3 
to 5. However in higher quality ratings wines achohol values tend to be higher.

# Bivariate Analysis

Most median and mean values from attributes doesn't vary alot or stays the same
in all ratings. However, alcohol, volatile.acidity, citric.acid, sulpahtes
stands out from all atributes by having most changes accross quality ratings. 

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?

Alcohol has the strongest relationship.

# Multivariate Plots Section

We will try to group all quality ratings in three groups:
 - Low rating group (ratings in range 3 to 5)
 - Midle rating group (ratings in range 5 to 6)
 - High rating group (ratings in range 6 to 8)

The reason of middle rating group to be "narrow" is because most of wines
gets ratings between 5 and 6.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create new column with quality category values
pf$quality_range.bucket <- cut(pf$quality, c(2,5,6,8),
                               labels = c("Low (3 to 5)",
                                          "Medium (6)",
                                          "High (7 to 8)"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Show alcohol and sulphates scatter plot with colors representing wine quality
# group.
ggplot(aes(x=alcohol, y=sulphates), data=pf) +
  geom_point(aes(color=quality_range.bucket)) + 
  scale_y_log10() + 
  scale_color_brewer("Quality\nCategories", type = 'div')
```

As wecan see, green dots shifted to the right compared yellow ones. That shows
wines that has higher quality rating might tend to have higher alcohol quantity.
Also, higher quality wines tend to have slightly higher sulphates quantities.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Show alcohol and volatile.axidity scatter plot with colors representing wine
# quality group.
ggplot(aes(x=alcohol, y=volatile.acidity), data=pf)+ geom_point(aes(color=quality_range.bucket)) + 
  scale_y_log10() + 
  scale_color_brewer("Quality\nCategories", type = 'div')
```

This graph indicates that lower rating group tend to have higher violatile
acidity and lower alcohol rating comapred to higher rating wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Show alcohol and citric.acid scatter plot with colors representing wine
# quality group.
ggplot(aes(x=alcohol, y=citric.acid), data=pf)+ 
  scale_y_log10() + 
  geom_point(aes(color=quality_range.bucket)) + 
  scale_color_brewer("Quality\nCategories", type = 'div')
```

Here we can see that "lower quality" wines tend to have low and high values of
citric acidity. However, "higher quality" wines a tend to have slighty more 
values in higher values of citric acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# New column with calculated quality_index based on alcohol citric.acid and
# volatile.acidity values
pf$quality_index <- (pf$alcohol * pf$citric.acid)/(pf$volatile.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatterplot (jittered) - Quality and Quality Index, by Quality Categories
ggplot(aes(x=quality, y=quality_index), data=pf) +
  geom_jitter(width = 0.35, height = 0, aes(color=quality_range.bucket)) + 
  scale_y_log10() + 
  scale_color_brewer("Quality\nCategories", type = 'div')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create linear model for quality rating and alcohol values.
m1 <- lm(I(quality) ~ I(alcohol), data = pf)
m2 <- update(m1, ~ . + volatile.acidity) # update model with volatile.acidity
m3 <- update(m2, ~ . + citric.acid) # update model with citric.acid values
m4 <- update(m3, ~ . + sulphates)# update model with sulphates values
mtable(m1, m2,m3, m4)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Features strengthen each, but not by much. Also alcohol and citric acidity has
positive relationship whereas violatile.acidity - negative relationship.

### Were there any interesting or surprising interactions between features?

The interesting findings is that how small affect has each of theses atributes.
Anotehr suprising thing - looks like strongest correlation has an alcohol 
compared to ohter atributes.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
Yes, we did linear regression with variables that looks like have most effect.
However results shows none of variables has an effect on wine ratings.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Print Quality rating distribution
ggplot(aes(x=quality), data=pf) +
  geom_histogram(color = 'black', binwidth = 1) +
  scale_x_continuous("Quality rating", breaks = seq(1,10,1)) +
  scale_y_continuous("Count") +
  ggtitle('Quality rating distribution') +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description One
As we cans see ratings are distributed almost in perfect noraml distribution. 
But even though ratings can be from 0 to 10, in real word it varies from 3 to 8.
Also there are no float ratings, because wines are evaluated in teger values.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Plot alcohol and quality rating scatter plot with mean and median lines.
ggplot(aes(x=quality, y=alcohol), data=pf) +
  geom_point(color = 'blue', alpha = 0.25, size = 1, position = 'jitter') +
  geom_line(aes(color = 'Mean'), stat = 'summary', fun.y = median) +
  geom_line(aes(color = 'Median'), stat = 'summary', linetype = "dashed",
            fun.y = mean) +
  scale_x_continuous("Quality rating") +
  scale_y_continuous("Alcohol (% by Volume)") +
  ggtitle('Wine quality by citric alcohol') +
  scale_color_manual('', values = c("Mean" = "green", "Median" = "red")) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Two
Looks like alcohol has a biggest effect on overall rating. Hewever it seems that
mean and median of alcohol values in quality ratings 3 to 5 remains almost
stable. But from rating 5, alcohol mean and median tends to rise. It seems that
on higher quality ratings, lowest alcohol values tend to grow up whereas highest
alcohol values remains the same acrross all ratings that are 5 or higher. 

### Plot Three
```{r echo=FALSE, Plot_Three}
# scatterplot (jittered) - Quality and Quality Index, by Quality Categories
ggplot(aes(x=quality, y=quality_index), data=pf) +
  geom_jitter(width = 0.35, height = 0, aes(color=quality_range.bucket)) + 
  scale_y_log10("Quality index") + 
  scale_color_brewer("Quality\nCategories", type = 'div') +
  ggtitle('Quality index comparison with quality values') +
  scale_x_continuous("Quality rating") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Three
In the last chart we compare Quality index with quality ratings. In perfect word
with every higher quality range (ratings 3-5, 5-6, 6-8), all wine should be in
higher position in the graph due to higher quality index. However this is not 
the case adn perhaps that's why the correlation between wine atributes and
quality ratings are low. As we can see lowest quality index in all cathegories
are the same through low middle and high rating groups. But highest quality 
indexes has small difference: low quality ratign group tend to have lower values
whereas middle and lower high rating wines tend to have the highest quality 
index values. What's interesting is the highest quality rating wines tend have 
lower values - all most the same as of quality rating 4.
------
  
# Reflection  
This tidy data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent). During this 
analysis I tried to look at histograms of every variable in order to get sense 
how values are distributed and how far they fluctuate. Then I compared every 
chemical variable with quality rating with that. The goal was to identify 
patterns that might affect rating values. After variables being  chosen, i check
how much effect they have in linear model as well as making plots with quality
rating groups.
  
Before starting this analysis I was hoping to find clear patterns of some 
variables affecting quality ratings. After seeing some histograms it was clear 
that some variables like density, pH has very low fluctuation. Also they are a 
byproduct of other variables (Density changes when fluid chemistry differs; pH 
changes when we have different amounts of acid or base elements. A lot of other 
variables are either acid or basic). When comparing quality rating with each 
every variable, there wasn’t any clear tendency. Most promising variable were 
alcohol, volatile acidity, citric acid and sulfates. However, when analyzing 
further with quality rating categories, comparing quality rating with quality 
index (calculated from chosen variables), patterns that chosen variables would 
affect quality ratings were minimal, if any. Lastly linear model supported my
findings: correlation was low. The findings were disappointing since I was 
expecting different results.
  
As any analysis it has some flaws that might influence wrong results. This might
be due to these reasons:
 - Because wines were evaluated by humans, rating could be influenced by subject 
feelings and tastes which this data not taking into account. Also psychological 
state of the members during testing can also make rating subjective.  
  
 1. There might be other attributes that aren't reflected in the data set. For
example color of wine, taste preferences of members of wine testing commission.  
 2. Small sample data. The lack of data was clear especially on low and high 
quality ratings;  
 3. All wines were made by 1 company. Diversity of wine suppliers would make
 data applicable about all red wine market.
It would be nice to expand this analysis with bigger, more diverse data set.
  
   
In general I’ve found this data analysis expand my view on wines. It also 
sharpen my data analysis skills and R language programming skills. Although it 
wasn’t easy, it was worth it.